Halving the cost of quantum addition

### 2018-04-04 paper.tex
Typo fixes and polish in opportunity cost section. Still one citation missing.

One of the downsides of the surface code is that it has no cheap mechanism to apply non-Clifford operations such as $T$ gates.
Instead, $T$ gates are performed by distilling and consuming $|T\rangle = \frac{1}{\sqrt{2}} (|0\rangle + e^{i \pi/4} |1\rangle)$ states.
The T gate is performed via a controlled-NOT from the target qubit into the $|T\rangle$ ancilla, followed by a classical feedback step that measures the ancilla to determine if an $S$ gate is applied to the target.
Unfortunately, distilling the $|T\rangle$ state that will ultimately be passed into this process has significant cost.
The cost of distillation is high enough that error corrected quantum computations are expected to bottleneck waiting for $|T\rangle$ states to be produced, so that their runtime is dominated by the number of T gates (as opposed to being dominated by gate count, circuit depth, or even measurement depth).
For example, although our adder has the same measurement depth as the Cuccaro adder (i.e. $2n + O(1)$), we still expect our adder to execute more quickly in practice on early error corrected machines (assuming there is room for the ancillae).

@@

Roughly speaking, state-of-the-art T factory designs that have been rigorously validated use 1000 units of ``spacetime volume" to produce one $|T\rangle$ state {\color{red} CITE CITE CITE}.
On the other hand, if an ancilla qubit is stored as a double-defect and lasts through $k$ serial measurements, this consumes at least $2k$ units of spacetime volume.
Putting these two facts together, and ignoring the fact that T factories are discrete objects with minimum sizes and packing constraints, we can say that the opportunity cost of holding an ancilla is one $|T\rangle$ state per 500 serial measurements.

With the above in mind, the ``effective" T-count of our adder, including the opportunity cost of ancillae, is $\frac{1}{500} n^2 + 4n + O(1)$ whereas the effective T-count of the Cuccaro adder is $8n + O(1)$.
For a sufficiently large addition, there is a cutoff point where ancillae are kept so long that the opportunity cost is larger than the four $|T\rangle$ states saved by each ancilla.
Beyond this cutoff, it is better to propagate carries inplace (as in the Cuccarro adder) instead of via ancillae.
Given the rough numbers we've given, the cutoff is approximately $n=2000$.
(Note that it is not necessary for the entire adder to use the same carry propagation strategy.
Parts below the cutoff can propagate carries via ancillae while parts above the cutoff propagate carries inplace.)

We caution that the cutoff we just estimated is not a fixed constant.
It is affected by other optimizations, by future research, and by overall system design.
As the spacetime volume of T factories is improved, the cutoff moves downward.
If the spacetime volume of qubits is improved, the cutoff moves upward.
For example, qubits that are idle can be compressed into an inactive ``memory" form that covers six times fewer physical qubits than the usual double-defect form {\color{red} CITE CITE CITE}.
This means that {\em idle} ancillae, like the ones in our adder, have a lower opportunity cost than active ancillae.
This optimization extends the cutoff from $n=2000$ to $n=12000$, beyond the size of adders that would be used to break modern RSA keys with Shor's algorithm.

An interesting consequence of ancillae having an T-count opportunity cost, besides making circuits that execute inplace desirable, is that it makes circuit depth relevant to the T-count.
Logarithmic-depth adders may use a constant factor more $|T\rangle$ states and ancillae than our adder, but they use exponentially less {\em ancilla-depth}.
The T-count opportunity cost of logarithmic-depth adders grows like $\Theta(n \lg n)$ instead of like $\Theta(n^2)$.
So, despite their ``raw" T-count being larger, for sufficiently large $n$ their effective T-count must become lower than the effective T-count of our ripple-carry adder.

We stress that the previous paragraph only applies in the regime where there are enough physical qubits to support enough T factories to properly feed a logarithmic-depth adder.
In the early error corrected regime, when only a few T factories are available, the constant-factor penalty on the T-count of logarithmic-depth adders will result in a runtime that is longer than the runtime of ripple-carry adders (because they both have to wait for $|T\rangle$ states).
Eventually, as more and more T factories become available, the parallelism inherent to logarithmic-depth adders will become accessible and overcome the constant-factor T-count penalty, resulting in a transition away from rippe-carry adders.

Whenever ancillae are used to save T gates, it is important to consider that there may be alternative uses of those ancillae that net even more $|T\rangle$ states.
These tradeoffs depend on overall system design.
Overall system optimization is important, but it is not the subject of this paper and so we will limit ourselves to introducing basic tools (e.g. the concept of a temporary logical-AND, as well as an adder that uses $4$ T gates and one ancillae per bit instead of $8$ T gates per bit) that future system architects can combine with other techniques on a case by case basis.

@@

Out-of-place adders are useful when a circuit is going to compute an addition, use it for awhile, then uncompute it.
Our out-of-place adder does not improve on the cost of computing an out-of-place addition.
However, because our out-of-place adder is based on computating a temporary logical-AND, its inverse does not use any T gates.
This makes it possible to uncompute an out-of-place addition while consuming no T gates.

@@

Yet another example of a kind of circuit that benefits from the temporary logical-AND is {\em low-depth} adder circuits.
For example, the Toffoli gates produced by the $P$ and $P^{-1}$ rounds in Draper et al's logarithmic-depth adder \cite{Draper2004} form compute/uncompute pairs that can be replaced by temporary logical-ANDs.
Alternatively, by using temporary logical-ANDs to implement the classical Brent-Kung adder \cite{brent1982} in a reversible fashion, two $N$-bit numbers can be added in $O(\lg N)$ depth with a T-count of only $12N$.

>>>>

One of the downsides of the surface code is that it has no cheap mechanism to apply non-Clifford operations such as $T$ gates.
Instead, $T$ gates are performed by distilling and consuming $|T\rangle = \frac{1}{\sqrt{2}} (|0\rangle + e^{i \pi/4} |1\rangle)$ states.
The T gate is performed via a controlled-NOT from the target qubit into the $|T\rangle$ ancilla, followed by a classical feedback step that measures the ancilla to determine if an $S$ gate is applied to the target.
Unfortunately, distilling the $|T\rangle$ state that will ultimately be passed into this process has significant cost.
The cost of distillation is high enough that error corrected quantum computations are expected to bottleneck waiting for $|T\rangle$ states to be produced, so that their runtime is dominated by the number of T gates (as opposed to being dominated by gate count, circuit depth, or even measurement depth).
For example, although our adder has the same measurement depth as the Cuccaro adder (i.e. $2n + O(1)$), we still expect our adder to execute more quickly in practice on early error corrected machines (assuming there is room for the $n$ ancillae our adder requires).

@@

It is possible to produce a high-quality $|T\rangle$ state in 960 units of ``spacetime volume" \cite{} (though this is certainly not a lower bound!).
For comparison, an ancilla qubit stored as a double-defect through $k$ serial measurements will cover at least $2k$ units of spacetime volume.
Putting these two facts together, and ignoring the fact that T factories are discrete objects with packing constraints, we will approximate the opportunity cost of holding an ancilla as $\frac{1}{480} |T\rangle$ states per measurement-depth.

With the above in mind, the ``effective" T-count of our adder, including the opportunity cost of ancillae (and accounting for the fact that some are kept for longer than others), is $\frac{1}{480} n^2 + 4n$ whereas the effective T-count of the Cuccaro adder is $8n$.
This implies that our adder is worse than the Cucarro adder when $n > 1920$.
However, switching adders at $n=1920$ ignores the possibility of using a hybrid of both adders where bit positions below a cutoff propagate carries via ancillae as in our adder and bit positions above the cutoff propagate carries inline as in the Cuccaro adder.
In a hybrid adder of this type, the cutoff actually occurs at $n=960$.

We caution that the cutoff of 960 that we just estimated is not a fixed constant.
It is affected by other optimizations, by future research, and by overall system design.
As the spacetime volume of T factories is improved, the cutoff {\em will} move downward.
Conversely, if the spacetime volume of ancillae is improved, the cutoff moves upward.
For example, qubits that are idle can be compressed into an inactive ``memory" form that covers six times fewer physical qubits than the double-defect form \cite{horsman2012}.
This means that {\em idle} ancillae, like the ones in our adder, have a lower opportunity cost than active ancillae.
This optimization increases the cutoff to $n \approx 5760$, which is beyond the size of adders that would be used to break modern RSA keys with Shor's algorithm.

An interesting consequence of ancillae having a T-count opportunity cost, besides making circuits that execute inplace desirable, is that it makes circuit depth relevant to the T-count.
Logarithmic-depth adders may use a constant factor more $|T\rangle$ states and ancillae than our adder, but they use exponentially less {\em ancilla-depth}.
The T-count opportunity cost of logarithmic-depth adders grows like $\Theta(n \lg n)$ instead of like $\Theta(n^2)$.
So, despite their ``raw" T-count being larger, for sufficiently large $n$ their effective T-count must become lower than the effective T-count of our ripple-carry adder.

Beware that the previous paragraph only applies in the regime where there are enough physical qubits to support enough T factories to properly feed a logarithmic-depth adder.
In the early error corrected regime, when only a few T factories are available, the constant-factor penalty on the T-count of logarithmic-depth adders will result in a runtime that is longer than the runtime of ripple-carry adders (because they both have to wait for $|T\rangle$ states).
It is only as more and more T factories become available that the parallelism inherent to logarithmic-depth adders becomes accessible, overcomes the constant-factor T-count penalty, and triggers a gradual transition away from ripple-carry adders.

Whenever ancillae are used to save T gates, it is important to consider that there may be alternative uses of those ancillae that net even more $|T\rangle$ states (or other desirable resources).
These tradeoffs depend on overall system design.
Overall system optimization is important, but it is not the subject of this paper and so we will limit ourselves to introducing basic tools (e.g. the concept of a temporary logical-AND, as well as an adder that uses $4$ T gates and one ancillae per bit instead of $8$ T gates per bit) that future system architects can combine with other techniques on a case by case basis.

@@

Out-of-place adders are useful when a circuit is going to compute an addition, use it for awhile, then uncompute it.
Our out-of-place adder does not improve on the cost of computing an out-of-place addition.
However, because our out-of-place adder is based on computing a temporary logical-AND, its inverse does not use any T gates.
This makes it possible to uncompute an out-of-place addition while consuming no T gates.

@@

Yet another example of a kind of circuit that benefits from the temporary logical-AND is {\em low-depth} adder circuits.
For example, the Toffoli gates produced by the $P$ and $P^{-1}$ rounds in Draper et al's logarithmic-depth adder \cite{Draper2004} form compute/uncompute pairs that can be replaced by temporary logical-ANDs.
Alternatively, by using temporary logical-ANDs to implement the classical Brent-Kung adder \cite{brent1982} in a reversible fashion, two $n$-bit numbers can be added in $O(\lg n)$ depth with a T-count of only $12n$.
